Modelling with ANIMO:\\ between Fuzzy Logic and Differential Equations

### 2015-06-26 Paper.tex
Updated model and data to the latest version, and improved the readme

\pgfdeclareimage[height=3ex]{win-logo}{images/win_logo}
\pgfdeclareimage[height=3ex]{mac-logo}{images/mac_logo}
\pgfdeclareimage[height=3ex]{linux-logo}{images/linux_logo}
\pgfdeclareimage[height=3ex]{ubuntu-logo}{images/ubuntu_logo}
\pgfdeclareimage[height=2ex]{mac-command}{images/mac_command}
\pgfdeclareimage[height=2ex]{lower-bound}{images/lowerbound}
\pgfdeclareimage[height=2ex]{upper-bound}{images/upperbound_green}
\pgfdeclareimage[width=0.95\textwidth]{100TNF0EGF}{images/TNFa-EGF/100TNFa-0EGF/100TNF_0EGF_ANIMOvsFuzzy-crop}
\pgfdeclareimage[width=0.95\textwidth]{0TNF100EGF}{images/TNFa-EGF/0TNFa-100EGF/0TNF_100EGF_ANIMOvsFuzzy-crop}
\pgfdeclareimage[width=0.95\textwidth]{100TNF100EGF}{images/TNFa-EGF/100TNFa-100EGF/100TNF_100EGF_ANIMOvsFuzzy-crop}
\pgfdeclareimage[width=0.95\textwidth]{CircadianComparison}{images/Circadian/ANIMOvsODE-crop}
\pgfdeclareimage[width=0.9\textwidth]{ANIMOvsODE-smaller-comparison}{images/Circadian/ANIMOvsODE_small_comparison-crop}

@@

\address[id=aff1]{%                           % unique id
  \orgname{Formal Methods and Tools, Faculty of EEMCS, University of Twente}, % university, etc
  %\street{Hallenweg},                     %
  \postcode{7522NH}                           % post or zip code
  \city{Enschede},                              % city
  \cny{The~Netherlands}                                    % country
}
\address[id=aff2]{%
  \orgname{Developmental BioEngineering, MIRA Institute for Biomedical Technology and Technical Medicine, University of Twente},
  %\street{ },
  \postcode{7522NH}
  \city{Enschede},
  \cny{The~Netherlands}
}
\address[id=aff3]{%
  \orgname{Human Media Interaction, Faculty of EEMCS, University of Twente},
  %\street{ },
  \postcode{7522NH}
  \city{Enschede},
  \cny{The~Netherlands}
}

@@

\begin{abstract} % abstract <= 350 words. Headers: background, results, conclusions
\parttitle{Background}
The large amount of data an experimental biologist
must deal with makes the use of computational methods essential
to reason on the dynamics of complex systems.
The software tool ANIMO (Analysis of Networks with Interactive MOdelling)
provides such computational support and allows insight into the complex networks of signalling events occurring in living cells.
Biology experts are able to exploit the power of the underlying formal methods
via a user interface specifically tailored for biological applications.
\parttitle{Results}
ANIMO models show a degree of precision intermediate between ordinary differential equations (ODEs) and fuzzy logic.
We show this by presenting ANIMO models for two case studies: \emph{Drosophila melanogaster} circadian clock, and signal
transduction events downstream of TNF$\alpha$ and EGF in HT-29 human
colon carcinoma cells.
The models were originally developed with ODEs and
fuzzy logic respectively.
\parttitle{Conclusions}
% The software tool ANIMO, recently updated to work with the latest version of Cytoscape,
% allows the biologist to develop a formal model through a user friendly interface,
% achieving useful insight without the need to study any mathematical formalism.
%
ANIMO models replicate with good precision the results of both the ODE and
fuzzy logic models. Moreover, ANIMO models require less parameters than ODEs and are more precise than
fuzzy logic. For this reason we position the modelling paradigm of ANIMO
between ODEs and fuzzy logic.
\end{abstract}

@@

\section*{Background}\label{sec:introduction}
\subsection*{Modelling in cell biology}
Executable biology is a young subfield in computational modelling, aimed at constructing models that mimic
biological phenomena \emph{in silico}. It provides an interesting paradigm for enhancing network
diagrams with an underlying formal description of both network components and their interactions.
Executable models typically contain descriptions of the direct interactions of network components,
requiring the modeller to think in terms of
``cause and effect''. Biologists reason in similar ways about the molecular mechanisms of network
interactions, and this makes the construction of an executable model an intuitive process.
ANIMO (Analysis of Networks with Interactive MOdelling,~\cite{animo-ieee,animo-gene})
provides an enabling technology to increase the use of computational models by experimental
biologists using their domain-specific language. Such models are indispensable for formally
comparing experimental data to prior knowledge, or for structuring experimental findings into a
new theory.
When dealing with complex biological networks executable biology models are particularly useful
to understand the non-linear dynamics and the entailed emergent properties of the networks.
In those cases, an ANIMO model can be used as a support to obtain insight based on abstract
representations of the interactions occurring inside living cells.
Other applications of ANIMO models include performing \emph{in silico} experiments and
the storage and transfer of knowledge on biological networks.

\subsection*{A brief introduction to ANIMO}
The user interface of ANIMO is displayed in Figure~\ref{fig:cytoscape},
where we present an example of a biological network enriched with dynamic information.
ANIMO is implemented as a plug-in to Cytoscape~\cite{cytoscape}, a software tool developed to represent biological networks.
On top of the static topological information displayed in Cytoscape, ANIMO represents
biological interactions starting from the basic concept of \emph{activity}:
each biological entity in an ANIMO model is considered to be either active or inactive.
Activity is to be intended in a very broad sense: for example, an active gene is being transcribed,
an active kinase can perform phosphorylations, etc. Each node in an ANIMO network represents
both active and inactive entities of the same type, with the relative amount of active entities
represented by the node colour on a user-configurable scale. Interactions among nodes define
how the biological entities in a network influence each other's activity.
Only nodes whose activity is larger than 0 can have an 
influence on their downstream targets, and only if that influence is not counterbalanced by intervening opposite interactions.
For example, the interaction $A\rightarrow B$ (read ``A activates B'') indicates that node A, if active,
will increase the activity of node B. If we add an additional interaction to the example,
$C \dashv$\hspace{0.1em}$B$ (``C inhibits B''), with C also active, then the activity of B will change depending on the
activity levels of A and C, and on the \emph{strength} of their influence.
The strength of an ANIMO interaction is a scale factor $k$ which defines the speed at which
that interaction occurs: higher values of $k$ give faster interactions. These $k$-values are
the only parameters needed in an ANIMO model, and can be given in either
a quantitative (as real numbers) or a qualitative way, choosing among self-explanatory descriptions
such as ``very slow'', ``slow'', ``medium'', ``fast'', ``very fast''.

\begin{figure*}[htbp]
\begin{center}
   \includegraphics[width=.85\textwidth]{images/screenshotTNFmodelSmall}
\end{center}
\caption{\csentence{The Cytoscape user interface running the ANIMO plug-in.}
The \emph{Network} panel in the centre contains the nodes-edges
model of the example TNF$\alpha$ pathway (see Methods Section), with
colours indicating node activity levels and shapes representing different protein categories (see the \emph{Legend} on the left).
The \emph{Results Panel} on the right contains a graph plotting activity levels of selected nodes
during the first 24 hours of evolution of the model. The slider under the graph
allows the user to select the time instant (marked as a vertical red line in the graph) on which
the colours nodes in the \emph{Network} are based.
The series with the {\sf \_data} suffix is experimental
data from~\cite{pathway-compendium}, considering a treatment with 100~ng/ml TNF$\alpha$.\\
All acronyms used in this paper
and their corresponding UniProt IDs are listed in Suppl. Sect.~\ref{suppl-sec:names}.
\label{fig:cytoscape}}
\end{figure*}

ANIMO models are analysed to produce graphs showing how the activity of selected nodes
change with time, allowing the user to obtain a view on the dynamic behaviour of their network.
In order to obtain these results, a model defined in ANIMO is automatically translated into
its corresponding representation as a network of Timed Automata~\cite{timed-automata-alur-dill}
and then analysed \emph{behind the scenes} with the software tool UPPAAL~\cite{uppaal}.
The formal language of Timed Automata allows to represent and analyse complex behaviours precisely and efficiently,
but the user does not need to directly manipulate Timed Automata or UPPAAL, as the analysis process is made transparent.
A fastidious user can still access the underlying models and perform other analyses in UPPAAL,
but that is not required in order to fully profit from ANIMO.

@@

\begin{figure*}[htbp]
 \begin{center}
  \includegraphics[width=0.9\textwidth]{images/modeling_methods_spectrum_animo}
 \end{center}
\caption{\csentence{The spectrum of modelling methods, with the addition of ANIMO.}
The precision of ANIMO models is halfway between fuzzy logic and ODE. In addition, ANIMO
allows for an easier modelling experience thanks to a user-friendly interface based on
the widely used network modelling software Cytoscape.\label{fig:animo-spectrum}}
\end{figure*}

@@

\section*{Results}\label{sec:results}
\subsection*{Modelling oscillation}\label{sec:animo-drosophila}
To demonstrate that results obtained with ANIMO are comparable to results from widely used modelling
approaches, we present 
an ANIMO model of the circadian clock in \emph{Drosophila Melanogaster} (Fig.~\ref{fig:drosophila-model}).
This ANIMO model is based on~\cite{drosophila-ode-model},
where ordinary differential equations (ODEs) were used.
The cyclic behaviour of the circadian clock is based on the alternating formation and destruction of the
CYC/CLK protein complex.
Concentration levels of this complex are in turn regulated by a series of proteins which are produced as
a consequence of CYC/CLK formation. The CWO protein
is central to the functioning of the network, as it degrades the mRNA for most of the involved proteins.
As such, CWO acts as an inhibitor that counterbalances the effect of CYC/CLK.
The positive influence of the light-regulated cryptochrome CRY on the degradation of TIM is a consequence
of the passage between day and night, allowing
the circadian clock to synchronize to a time zone (see Suppl. Sect.~\ref{suppl:repressilator}).

@@

\begin{figure*}[!htpb]
\begin{center}
\pgfuseimage{ANIMOvsODE-smaller-comparison}
\end{center}
\caption{\label{fig:grafici-drosophila-smaller}\csentence{Comparing the result of the ANIMO model of \emph{Drosophila Melanogaster} circadian clock
with the model of~\cite{drosophila-ode-model}.} 24 hours simulations of the two models were compared
against each other, synchronizing their start point as much as possible. The blue line is the ANIMO model
({\sf \_{}ANIMO} series), while the red line represents the data computed from the original ODE model
({\sf \_{}ODE} series). The vaules of all series were rescaled on a [0, 100] interval to facilitate comparison.}
\end{figure*}

\subsection*{Using ANIMO to generate hypotheses}\label{subsec:case-study-larger}
We now present a comparison with an existing fuzzy logic model, which we use also as an example of how
ANIMO can be used to create reference models and help to obtain insight into complex biological events.

@@

\def\largeModelScale{0.18}%0.15}%0.155}%0.27}
\def\legendScalaColori{0.21}%0.18}%0.21}
\def\legendScalaForme{0.21}%0.18}%0.21}
\def\scalaGrafici{0.0709}%0.13}
\begin{figure*}[!htpb]
\centering
%  \subfloat[\label{fig:large-model-no-hypotheses}]{\includegraphics[scale=\largeModelScale]{images/large_network_merged_no_hypotheses}}\\
% %\subfloat{\includegraphics[scale=\legendScalaColori]{images/legenda_forme}}\\ \subfloat{\includegraphics[scale=\legendScalaForme]{images/legenda_colori}}\\
% \subfloat{\includegraphics[scale=\legendScalaColori]{images/legenda_forme_e_colori}}\\
% \addtocounter{subfigure}{-1}
% \subfloat[\label{fig:large-model-graph1}]{\includegraphics[scale=\scalaGrafici]{images/TGF100_TNF0_ho_hyp_IL-1a3}}
% \subfloat[\label{fig:large-model-graph3}]{\includegraphics[scale=\scalaGrafici]{images/TNF5_C225_no_hyp_JNK1_MK2}}
  \includegraphics[width=0.9\textwidth]{images/large_network_merged_no_hypotheses_completa}
\caption{\csentence{Signalling network downstream of TNF$\alpha$ and EGF in human colon carcinoma cells.}
{\bf(A)} %{\bf \protect\subref{fig:large-model-no-hypotheses}}
The model for the merged TNF$\alpha$ and EGF pathways. Node colours represent the
activity level of the corresponding modelled reactants at time $t = 15$ minutes after
a stimulation of 100 ng/ml TNF$\alpha$ + 100 ng/ml EGF.
{\bf(B)} %{\bf \protect\subref{fig:large-model-graph1}}~
Modelled production of IL-1$\alpha$ after stimulation with 100 ng/ml TGF$\alpha$ (24 hours).
{\bf(C)} %{\bf \protect\subref{fig:large-model-graph3}}~
Modelled activation of JNK1 and MK2 after stimulation with 5 ng/ml TNF$\alpha$ + 10 $\mu$g/ml C225 (2 hours).
\\
The {\sf \_{}data} suffix identifies experimental data; all other series are computed by ANIMO.}\label{fig:large-model-all}
\end{figure*}

@@

\begin{figure*}[!htpb]
\begin{center}
% \subfloat[\label{fig:large-model-complete}]{\includegraphics[scale=\largeModelScale]{images/large_network_legendg}}\\
%  \subfloat[\label{fig:large-model-graph2}]{\includegraphics[scale=\scalaGrafici]{images/TGF100_vs_TNF100_hyp1_IL-1a4}}
%  \subfloat[\label{fig:large-model-graph4}]{\includegraphics[scale=\scalaGrafici]{images/TNF5_C225_hyp2_JNK1_MK2}}
  \includegraphics[width=0.9\textwidth]{images/large_network_legendg_completa}
\end{center}
\caption{\csentence{Signalling network downstream of TNF$\alpha$ and EGF in human colon carcinoma cells: improved model.}
{\bf(A)} %{\bf \protect\subref{fig:large-model-complete}}
The model for the merged TNF$\alpha$ and EGF pathways
after addition of the two hypotheses (highlighted).
Hypothesis 1 assumes IL-1$\alpha$ expression to depend on AP-1 activity, which in turn requires
both c-Jun en c-Fos to be activated by JNK1 and ERK, respectively. Hypothesis 2 assumes RAS as an activator
of MEKK1. Node colours represent the activity levels $15$ minutes
after stimulation of 100 ng/ml TNF$\alpha$ + 100 ng/ml EGF.
{\bf(B)} %{\bf \protect\subref{fig:large-model-graph2}}~
After the addition of the first hypothesis (activation of IL-1$\alpha$ production depending both
on JNK1 and ERK): production of IL-1$\alpha$ after stimulation with 100 ng/ml TNF$\alpha$ (series {\sf IL-1a~(TNFa)})
compared with stimulation with 100 ng/ml TGF$\alpha$ (series {\sf IL-1a~(TGFa)}) (24 hours).
{\bf(C)} %{\bf \protect\subref{fig:large-model-graph4}}~
After the addition of the second hypothesis (activation of MEKK1 downstream of EGFR):
stimulation with 5 ng/ml TNF$\alpha$ + 10 $\mu$g/ml C225 (2 hours).
Suppl. Sect.~\ref{suppl:parameters-tnf-egf} explains how the dosage of 5 ng/ml TNF$\alpha$ was represented in the model.\\
The {\sf \_{}data} suffix identifies experimental data; all other series are computed by ANIMO.}\label{fig:large-model-complete}
\end{figure*}

@@

\section*{Discussion}
\subsection*{Final remarks on the models}
We described the construction of an ANIMO
model of the circadian clock in \emph{Drosophila Melanogaster}. This model
captured the dynamics of the regulatory network and led to similar
conclusions as an ODE model that had been
published previously~\cite{drosophila-ode-model}. This finding supports the use of
the series of modelling abstractions on which ANIMO is based. The biggest
difference between the construction of these models is that the model in~\cite{drosophila-ode-model}
is constructed by writing a system of mathematical equations, together
with an algorithm for simulation. In ANIMO a number of network
nodes is drawn for the molecules involved.
These nodes are then linked by directed
interactions that represent cause-and-effect relationships, with a single parameter
that defines the strength of each
interaction. This approach to construct a model is closer to biological practice.
Further contributing to an interactive modelling process
is the compositionality of the model. Each node in the network
can be disabled at any time by the user, or extra nodes can be added,
without having to change any of the existing interactions.
All this comes at the price of lower model precision: the curves representing
oscillation of protein activities in the ANIMO model are not as smooth
nor as precise as those obtained from the original ODE model (see Suppl. Fig.~\ref{suppl-fig:grafici-drosophila}).

@@

All reactants modelled in ANIMO are assumed to be present at the same
(unitary) concentration. While this assumption is not always applicable, it encourages
abstract thought: many biological processes can be represented as networks driven by
\emph{activity-based} interactions (see Methods Section). Even if with a limited scope, ANIMO can be applied
also in the analysis of metabolic processes, considering the production of protein $A$ as
the activation of node $A$, with degradation being represented as inhibition. As stated before,
it cannot be expected from such models to be a completely realistic representation of their target biological
processes, but they can still be an useful tool. This can be seen for example in the circadian clock
model in the Results section, where mRNA and protein concentrations were abstracted to
activity-driven processes in the ANIMO model.

@@

\section*{Additional Files}
  \subsection*{Additional file 1 --- ANIMO models and data for the case studies}
      The provided .zip file contains the Cytoscape .cys network files which can be analysed
      with ANIMO. In addition, the reference values in .csv format are provided, to compare
      ANIMO's results with the previously existing models.

@@

\addtocontents{toc}{\protect\setcounter{tocdepth}{5}}
\thispagestyle{empty}
\tableofcontents
\clearpage

>>>>

\pgfdeclareimage[height=3ex]{win-logo}{images/win_logo}
\pgfdeclareimage[height=3ex]{mac-logo}{images/mac_logo}
\pgfdeclareimage[height=3ex]{linux-logo}{images/linux_logo}
\pgfdeclareimage[height=3ex]{ubuntu-logo}{images/ubuntu_logo}
\pgfdeclareimage[height=2ex]{mac-command}{images/mac_command}
\pgfdeclareimage[height=2ex]{lower-bound}{images/lowerbound}
\pgfdeclareimage[height=2ex]{upper-bound}{images/upperbound_green}
\pgfdeclareimage[width=0.95\textwidth]{FigS10}{Figures/S10} %100TNF0EGF}{images/TNFa-EGF/100TNFa-0EGF/100TNF_0EGF_ANIMOvsFuzzy-crop}
\pgfdeclareimage[width=0.95\textwidth]{FigS11}{Figures/S11} %0TNF100EGF}{images/TNFa-EGF/0TNFa-100EGF/0TNF_100EGF_ANIMOvsFuzzy-crop}
\pgfdeclareimage[width=0.95\textwidth]{FigS12}{Figures/S12} %100TNF100EGF}{images/TNFa-EGF/100TNFa-100EGF/100TNF_100EGF_ANIMOvsFuzzy-crop}
\pgfdeclareimage[width=0.95\textwidth]{FigS6}{Figures/S6} %CircadianComparison}{images/Circadian/ANIMOvsODE-crop}
\pgfdeclareimage[width=0.9\textwidth]{Fig3}{Figures/3} %ANIMOvsODE-smaller-comparison}{images/Circadian/ANIMOvsODE_small_comparison-crop}

@@

\address[id=aff1]{%                           % unique id
  \orgname{Formal Methods and Tools, Faculty of EEMCS, University of Twente}, % university, etc
  \street{P.O. Box 217},                     %
  \postcode{7500AE}                           % post or zip code
  \city{Enschede},                              % city
  \cny{The~Netherlands}                                    % country
}
\address[id=aff2]{%
  \orgname{Developmental BioEngineering, MIRA Institute for Biomedical Technology and Technical Medicine, University of Twente},
  \street{P.O. Box 217},                     %
  \postcode{7500AE}                           % post or zip code
  \city{Enschede},
  \cny{The~Netherlands}
}
\address[id=aff3]{%
  \orgname{Human Media Interaction, Faculty of EEMCS, University of Twente},
  \street{P.O. Box 217},                     %
  \postcode{7500AE}                           % post or zip code
  \city{Enschede},
  \cny{The~Netherlands}
}

@@

\begin{abstract} % abstract <= 350 words. Headers: background, results, conclusions
\parttitle{Background}
Experimental biologists need to deal with large amounts of data on a daily basis.
This makes computational support essential in order
to reason on the dynamics of biological systems.
% The large amount of data an experimental biologist
% must deal with makes the use of computational methods essential
% to reason on the dynamics of complex systems.
The software tool ANIMO (Analysis of Networks with Interactive MOdelling)
provides such computational support and allows insight into the complex networks of signalling events occurring in living cells.
Biology experts are able to exploit the power of the underlying formal methods
via a user interface specifically tailored for biological applications.
\parttitle{Results}
ANIMO models show a degree of precision intermediate between ordinary differential equations (ODEs) and fuzzy logic.
We show this by presenting ANIMO models for two case studies: \emph{Drosophila melanogaster} circadian clock, and signal
transduction events downstream of TNF$\alpha$ and EGF in HT-29 human
colon carcinoma cells.
The models were originally developed with ODEs and
fuzzy logic respectively.
\parttitle{Conclusions}
% The software tool ANIMO, recently updated to work with the latest version of Cytoscape,
% allows the biologist to develop a formal model through a user friendly interface,
% achieving useful insight without the need to study any mathematical formalism.
%
ANIMO models replicate with good precision the results of both the ODE and
fuzzy logic models. Moreover, ANIMO models require less parameters than ODEs and are more precise than
fuzzy logic. For this reason we position the modelling paradigm of ANIMO
between ODEs and fuzzy logic.
\end{abstract}

@@

\section*{Background}\label{sec:introduction}
\subsection*{Modelling in cell biology}
Executable biology is a young subfield in computational modelling, aimed at constructing models that mimic
biological phenomena \emph{in silico}. It provides an interesting paradigm 
% that enriches the
% normally static biological network diagrams with dynamic information. An underlying formalism is used 
% to describe
% the dynamics of a network and its components. %, defining the executable model.
% The analysis of an executable model typically produces graphs and animations, which help understand how the network evolves with time.
% Executable models typically contain descriptions of the 
to enhance network diagrams with an underlying formal description of network components
and their interactions. Executable models typically contain descriptions of the 
direct interactions of network components, requiring the modeller to think in terms of
``cause and effect''. Biologists reason in similar ways about the molecular mechanisms of network
interactions, and this makes the construction of an executable model an intuitive process.
ANIMO (Analysis of Networks with Interactive MOdelling,~\cite{animo-ieee,animo-gene})
provides an enabling technology to increase the use of computational models by experimental
biologists using their domain-specific language. 
ANIMO enriches the normally static biological network diagrams with dynamic information,
which is then used to automatically produce formal models representing the biological network.
Such models are indispensable for formally
comparing experimental data with prior knowledge, or for structuring experimental findings into a
new theory.
When dealing with complex biological networks, executable biology models are particularly useful
to understand the non-linear dynamics and the entailed emergent properties of the networks.
In those cases, an ANIMO model can be used as a support to obtain insight based on abstract
representations of the interactions occurring inside living cells.
Other applications of ANIMO models include performing \emph{in silico} experiments and
the storage and transfer of knowledge on biological networks.

\subsection*{A brief introduction to ANIMO}
The user interface of ANIMO is displayed in Figure~\ref{fig:cytoscape},
where we present an example of a biological network enriched with dynamic information.
ANIMO is implemented as a plug-in to Cytoscape~\cite{cytoscape}, a software tool developed to represent biological networks.
On top of the static topological information displayed in Cytoscape, ANIMO represents
biological interactions starting from the basic concept of \emph{activity}:
each biological entity in an ANIMO model is considered to be either active or inactive.
Activity is to be interpreted in a very broad sense: for example, an active gene is being transcribed,
an active kinase can perform phosphorylations, etc. Each node in an ANIMO network represents
both active and inactive entities of the same type, with the relative amount of active entities
(the \emph{activity level}) represented by the node colour on a user-configurable scale.
Interactions among nodes define how the biological entities in a network influence each other's activity.
Only nodes whose activity level is larger than 0 (such nodes are called \emph{active}) can have an 
influence on their downstream targets, and only if that influence is not counterbalanced by intervening opposite interactions.
For example, the interaction $A\rightarrow B$ (read ``A activates B'') indicates that node A, if active,
will increase the activity of node B. If we add an additional interaction to the example,
$C \dashv$\hspace{0.1em}$B$ (``C inhibits B''), with C also active, then the activity level of B will change depending on the
activity levels of A and C, and on their quantitative influence.
The influence of an ANIMO interaction is quantified by a scale factor $k$, which defines the speed at which
that interaction occurs: higher values of $k$ give faster interactions. These $k$-values are
the only parameters needed in an ANIMO model, and can be given in either
a quantitative (as real numbers) or a qualitative way, choosing among self-explanatory descriptions
such as ``very slow'', ``slow'', ``medium'', ``fast'', ``very fast''.

\begin{figure*}[htbp]
\begin{center}
%    \includegraphics[width=.85\textwidth]{images/screenshotTNFmodelSmall}
  \includegraphics[width=.85\textwidth]{Figures/1}
\end{center}
\caption{\csentence{The Cytoscape user interface running the ANIMO plug-in.}
The \emph{Network} panel in the centre contains the nodes-edges
model of the example TNF$\alpha$ pathway (see Methods Section), with
colours indicating node activity levels and shapes representing different protein categories (see the \emph{Legend} on the left).
The \emph{Results Panel} on the right contains a graph plotting activity levels of selected nodes
during the first 24 hours of simulation of the model. The slider under the graph
allows the user to select the time instant (marked as a vertical red line in the graph) on which
the colours nodes in the \emph{Network} are based.
The series with the {\sf \_data} suffix is experimental
data from~\cite{pathway-compendium}, considering a treatment with 100~ng/ml TNF$\alpha$.\\
All acronyms used in this paper
and their corresponding UniProt IDs are listed in Suppl. Sect.~\ref{suppl-sec:names}.
\label{fig:cytoscape}}
\end{figure*}

ANIMO produces graphs showing how the activity levels of selected nodes
change over time, allowing the user to obtain a view on the dynamic behaviour of their network.
In order to obtain these results, a model defined in ANIMO is automatically translated into
its corresponding representation as a network of Timed Automata~\cite{timed-automata-alur-dill}
and then analysed \emph{behind the scenes} with the software tool UPPAAL~\cite{uppaal}.
The formal language of Timed Automata allows to represent and analyse complex behaviours precisely and efficiently,
but the user does not need to directly manipulate Timed Automata or UPPAAL, as the analysis process is made transparent.
A fastidious user can still access the underlying models and perform other analyses in UPPAAL,
but that is not required in order to fully profit from ANIMO.

@@

\begin{figure*}[htbp]
 \begin{center}
%   \includegraphics[width=0.9\textwidth]{images/modeling_methods_spectrum_animo}
  \includegraphics[width=0.9\textwidth]{Figures/2}
 \end{center}
\caption{\csentence{The spectrum of modelling methods, with the addition of ANIMO.}
The precision of ANIMO models is halfway between fuzzy logic and ODE. In addition, ANIMO
allows for an easier modelling experience thanks to a user-friendly interface based on
the widely used network modelling software Cytoscape.\label{fig:animo-spectrum}}
\end{figure*}

@@

\section*{Results}\label{sec:results}
\subsection*{Modelling oscillation in Drosophila Melanogaster circadian clock}\label{sec:animo-drosophila}
To demonstrate that results obtained with ANIMO are comparable to results from widely used modelling
approaches, we present 
an ANIMO model of the circadian clock in \emph{Drosophila Melanogaster} (Fig.~\ref{fig:drosophila-model}).
This ANIMO model is based on~\cite{drosophila-ode-model},
where ordinary differential equations (ODEs) were used.
The cyclic behaviour of the circadian clock is based on the alternating formation and destruction of the
CYC/CLK protein complex.
Concentration levels of this complex are in turn regulated by a series of proteins which are produced as
a consequence of CYC/CLK formation. The CWO protein
is central to the functioning of the network, as it degrades the mRNA for most of the involved proteins.
As such, CWO acts as an inhibitor that counterbalances the effect of CYC/CLK.
The positive influence of the light-regulated cryptochrome CRY on the degradation of TIM is a consequence
of the passage between day and night, allowing
the circadian clock to synchronize to a time zone (see Suppl. Sect.~\ref{suppl:repressilator}).

@@

\begin{figure*}[!htpb]
\begin{center}
% \pgfuseimage{ANIMOvsODE-smaller-comparison}
  \pgfuseimage{Fig3}
\end{center}
\caption{\label{fig:grafici-drosophila-smaller}\csentence{Comparing the result of the ANIMO model of \emph{Drosophila Melanogaster} circadian clock
with the model of~\cite{drosophila-ode-model}.} 24 hours simulations of the two models were compared
against each other, synchronizing their start point as much as possible. The blue line is the ANIMO model
({\sf \_{}ANIMO} series), while the red line represents the data computed from the original ODE model
({\sf \_{}ODE} series). The vaules of all series were rescaled on a [0, 100] interval to facilitate comparison.}
\end{figure*}

\subsection*{Using ANIMO to generate hypotheses in human colon carcinoma cells}\label{subsec:case-study-larger}
We now present a comparison with an existing fuzzy logic model, which we use also as an example of how
ANIMO can be used to create reference models and help to obtain insight into complex biological events.

@@

\def\largeModelScale{0.18}%0.15}%0.155}%0.27}
\def\legendScalaColori{0.21}%0.18}%0.21}
\def\legendScalaForme{0.21}%0.18}%0.21}
\def\scalaGrafici{0.0709}%0.13}
\begin{figure*}[!htpb]
\centering
%  \subfloat[\label{fig:large-model-no-hypotheses}]{\includegraphics[scale=\largeModelScale]{images/large_network_merged_no_hypotheses}}\\
% %\subfloat{\includegraphics[scale=\legendScalaColori]{images/legenda_forme}}\\ \subfloat{\includegraphics[scale=\legendScalaForme]{images/legenda_colori}}\\
% \subfloat{\includegraphics[scale=\legendScalaColori]{images/legenda_forme_e_colori}}\\
% \addtocounter{subfigure}{-1}
% \subfloat[\label{fig:large-model-graph1}]{\includegraphics[scale=\scalaGrafici]{images/TGF100_TNF0_ho_hyp_IL-1a3}}
% \subfloat[\label{fig:large-model-graph3}]{\includegraphics[scale=\scalaGrafici]{images/TNF5_C225_no_hyp_JNK1_MK2}}
%   \includegraphics[width=0.9\textwidth]{images/large_network_merged_no_hypotheses_completa}
  \includegraphics[width=0.9\textwidth]{Figures/4}
\caption{\csentence{Signalling network downstream of TNF$\alpha$ and EGF in human colon carcinoma cells.}
{\bf(A)} %{\bf \protect\subref{fig:large-model-no-hypotheses}}
The model for the merged TNF$\alpha$ and EGF pathways. Node colours represent the
activity level of the corresponding modelled reactants at time $t = 15$ minutes after
a stimulation of 100 ng/ml TNF$\alpha$ + 100 ng/ml EGF.
{\bf(B)} %{\bf \protect\subref{fig:large-model-graph1}}~
Modelled production of IL-1$\alpha$ after stimulation with 100 ng/ml TGF$\alpha$ (24 hours).
{\bf(C)} %{\bf \protect\subref{fig:large-model-graph3}}~
Modelled activation of JNK1 and MK2 after stimulation with 5 ng/ml TNF$\alpha$ + 10 $\mu$g/ml C225 (2 hours).
\\
The {\sf \_{}data} suffix identifies experimental data; all other series are computed by ANIMO.}\label{fig:large-model-all}
\end{figure*}

@@

\begin{figure*}[!htpb]
\begin{center}
% \subfloat[\label{fig:large-model-complete}]{\includegraphics[scale=\largeModelScale]{images/large_network_legendg}}\\
%  \subfloat[\label{fig:large-model-graph2}]{\includegraphics[scale=\scalaGrafici]{images/TGF100_vs_TNF100_hyp1_IL-1a4}}
%  \subfloat[\label{fig:large-model-graph4}]{\includegraphics[scale=\scalaGrafici]{images/TNF5_C225_hyp2_JNK1_MK2}}
%  \includegraphics[width=0.9\textwidth]{images/large_network_legendg_completa}
  \includegraphics[width=0.9\textwidth]{Figures/5}
\end{center}
\caption{\csentence{Signalling network downstream of TNF$\alpha$ and EGF in human colon carcinoma cells: improved model.}
{\bf(A)} %{\bf \protect\subref{fig:large-model-complete}}
The model for the merged TNF$\alpha$ and EGF pathways
after addition of the two hypotheses (highlighted).
Hypothesis 1 assumes IL-1$\alpha$ expression to depend on AP-1 activity, which in turn requires
both c-Jun en c-Fos to be activated by JNK1 and ERK, respectively. Hypothesis 2 assumes RAS as an activator
of MEKK1. Node colours represent the activity levels $15$ minutes
after stimulation of 100 ng/ml TNF$\alpha$ + 100 ng/ml EGF.
{\bf(B)} %{\bf \protect\subref{fig:large-model-graph2}}~
After the addition of the first hypothesis (activation of IL-1$\alpha$ production depending both
on JNK1 and ERK): production of IL-1$\alpha$ after stimulation with 100 ng/ml TNF$\alpha$ (series {\sf IL-1a~(TNFa)})
compared with stimulation with 100 ng/ml TGF$\alpha$ (series {\sf IL-1a~(TGFa)}) (24 hours).
{\bf(C)} %{\bf \protect\subref{fig:large-model-graph4}}~
After the addition of the second hypothesis (activation of MEKK1 downstream of EGFR):
stimulation with 5 ng/ml TNF$\alpha$ + 10 $\mu$g/ml C225 (2 hours).
Suppl. Sect.~\ref{suppl:parameters-tnf-egf} explains how the dosage of 5 ng/ml TNF$\alpha$ was represented in the model.\\
The {\sf \_{}data} suffix identifies experimental data; all other series are computed by ANIMO.}\label{fig:large-model-complete}
\end{figure*}

@@

\section*{Discussion}
\subsection*{Final remarks on the models}
We first described the construction of an ANIMO
model of the circadian clock in \emph{Drosophila Melanogaster}. This model
captured the dynamics of the regulatory network and led to similar
conclusions as an ODE model that had been
published previously~\cite{drosophila-ode-model}. This finding supports the use of
the series of modelling abstractions on which ANIMO is based. The biggest
difference between the construction of these models is that the model in~\cite{drosophila-ode-model}
is constructed by writing a system of mathematical equations, together
with an algorithm for simulation. In ANIMO a number of network
nodes is drawn for the molecules involved.
These nodes are then linked by directed
interactions that represent cause-and-effect relationships, with a single parameter
that defines the quantitative effect of each
interaction. This approach to construct a model is closer to biological practice.
Further contributing to an interactive modelling process
is the compositionality of the model. Each node in the network
can be disabled at any time by the user, or extra nodes can be added,
without having to change any of the existing interactions.
All this comes at the price of lower model precision: the curves representing
oscillation of protein activities in the ANIMO model are not as smooth
nor as precise as those obtained from the original ODE model (see Suppl. Fig.~\ref{suppl-fig:grafici-drosophila}).

@@

For each reactant modelled in ANIMO we assume that the total amount
of active and inactive molecules remains constant.
While this assumption is not always applicable, it encourages
abstract thought: many biological processes can be represented as networks driven by
\emph{activity-based} interactions (see Methods Section). Even if with a limited scope, ANIMO can be applied
also in the analysis of metabolic processes, considering the production of protein $A$ as
the activation of node $A$, with degradation being represented as inhibition. As stated before,
it cannot be expected from such models to be a completely realistic representation of their target biological
processes, but they can still be a useful tool. This can be seen for example in the circadian clock
model in the Results section, where mRNA and protein concentrations were abstracted to
activity-driven processes in the ANIMO model.

@@

\section*{Additional Files}
  \subsection*{Additional file 1 --- ANIMO models and data for the case studies}
      The provided .zip file contains the Cytoscape .cys network files which can be analysed
      with ANIMO. In addition, the reference values in .csv format are provided, to compare
      ANIMO's results with the previously existing models. In order to grant a fair comparison,
      note that ANIMO's results need to be rescaled the same way as the ODE results
      (i.e. normalize each series w.r.t. to its minimum and maximum) before the comparison.

@@

\startcontents[appendices]
\stopcontents[appendices]
